Relational Parametricity for Algebraically Indexed Types

### 2012-07-03 paper/section-further-examples.tex
Additions to the general framework, and fill in some of the examples
- Add a subsection on the construction of relational environments (unfinished)
- Fill in the monoid indexed types section
- Rearrange the example sections a bit

\section{More Algebraically Indexed Types}
\label{sec:further-examples}

\subsection{Singleton Types}
\label{sec:singleton-types}

@@

\begin{enumerate}
\item Take integers as a useful example
\item The relational interpretation is simply that two things are
  related if they are both equal to the indexing value (and each
  other).
\item Types are extremely precise... Can this be quantified?
\item Indefinability result by considering sub-algebras? Is this a
  general method? Can't write:
  \begin{displaymath}
    \forall z_1,z_2,z_3.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to \tyPrim{int}{z_3}
  \end{displaymath}
  because there is no way of getting $z_3$ from $z_1$ and $z_2$. Can
  prove this using the relational environments. Is there a nice
  matrix-style way of characterising the allowable things? I guess it
  is harder when there are two operations lying around (multiplication
  and addition here). Maybe we could restrict to multiplication by
  constants? Then we could possibly characterise the definable
  elements by a matrix representation?
\end{enumerate}

\subsection{Predicate Types}
\label{sec:predicate-types}

\begin{enumerate}
\item Let the indexing theory be predicates on the underlying value
\item With the laws of boolean algebra
\item Singleton types are a special case (i.e. with equality to a fixed value)
\item But also have $\land$, $\lor$, $\lnot$ (and derived $\Rightarrow$)
\item A pair of things is related if they both satisfy the predicate
  (and are equal?)
\item Need a special type for comparisons?
  \begin{displaymath}
    (<) : \forall z_1,z_2:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to (\tyPrim{int}{< z_1} \times 
  \end{displaymath}
\end{enumerate}

\subsection{Logical Information Flow}
\label{sec:information-flow}

@@

\subsection{Metric Spaces and Simple Continuity Analysis}
\label{sec:continuity-analysis}

@@

\subsection{Monoid Indexed Types}
\label{sec:monoid-indexed-types}

Primitive type is a monoid, and we index by a commutative monoid. An
example of this is strings with tainting information, or bags, or sets
of stuff. This gives a slightly different information flow analysis to
the logical information flow above. In particular, we can now count
uses of input data.

Primitive operations $\Gamma_{\mathit{Mon}}$:
\begin{displaymath}
  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
    1 & \tyPrim{M}{1} \\
    (\cdot) & \forall a,b.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a\cdot b}
  \end{array}
\end{displaymath}

For the index-erasure interpretation, we assume a monoid
$\mathcal{M}$. The primitive type $\tyPrimNm{M}$ is interpreted as
$\tySem{\tyPrim{M}{e}} = \mathcal{M}$. The primitive operations in
$\Gamma_{\mathit{Mon}}$ are interpreted just as the unit and
multiplication of the monoid $\mathcal{M}$.

% Need the following definitions: syntatic group/monoid arising from
% the syntax, and a neater way of defining relational
% environments. Also, a neat way of formally introducing a load of
% primitive operations.

For each index context $\Delta$, the set
$\relEnv{E}_\Mon(\Delta)$ of relational environments is
defined to be $\{ \rho_{(M,h)} \sepbar M\textrm{ is a submonoid of
}(\Delta \Rightarrow \mathsf{M})/\equiv, h\textrm{ is a monoid
  homomorphism }M \to Z(\mathcal{M}) \}$, where
\begin{displaymath}
  \rho_{(M,h)}\ \tyPrimNm{M}\ (e) = \left\{
    \begin{array}{ll}
      \{ \} & \textrm{if}\ e\not\in M \\
      \{ (x, h(e)\cdot x) \sepbar x \in \mathcal{M} \} & \textrm{if}\ e\in M
    \end{array}
  \right.
\end{displaymath}
Note the mapping to the centre of the monoid
$\mathcal{M}$. \fixme{introduce the centre notation earlier on}

\begin{lemma}
  For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
  $(\eta_\Mon,\eta_\Mon) \in \rsem{\Gamma_\Mon}{\relEnv{E}}\rho$.
\end{lemma}

\begin{theorem}
  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
  there exists a program $M$ such that
  \begin{equation}
    \label{eq:fo-monoid-type}
    \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
  \end{equation}
  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e =
  i_1^{b_1}\dots i_m^{b_m}$, if and only if there is a solution
  $\vec{x} = (x_1,\cdots,x_n)$ in the natural numbers to the equation
  $A x^\top = b$.
\end{theorem}

\begin{proof}
  (If) The program $\Lambda i_1\dots i_m.\ \lambda z_1\dots z_n.\
  z_1^{x_1}\dots z_n^{x_n}$ satisfies the typing judgment
  (\ref{eq:fo-monoid-type}).

  (Only if) Assume that there is a program $M$ satisfying the typing
  judgment (\ref{eq:fo-monoid-type}). By \thmref{thm:abstraction} we
  know that for all relational environments $\rho \in
  \relEnv{E}(i_1,\dots,i_m)$, and all $z_j,z'_j \in \mathcal{M}$,
  \begin{displaymath}
    \begin{array}{l}
      (z_1,z'_1) \in \rho\ \tyPrimNm{M}\ (e_1) \land \dots \land (z_n,z'_n) \in \rho\ \tyPrimNm{M}\ (e_n) \Rightarrow \\
      \quad (\tmSem{M}z_1\dots z_n, \tmSem{M}z'_1\dots z'_n) \in \rho\ \tyPrimNm{M}\ (e)
    \end{array}
  \end{displaymath}
  We select the relational environments
  \begin{displaymath}
    \rho\ \tyPrimNm{M}\ (e) = \left\{
      \begin{array}{ll}
        \{(z,z) \sepbar z \in \mathcal{M} \} &
        \begin{array}[t]{@{}l}
          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
          \textrm{ for some }x_1,\dots,x_n \in \mathbb{N}
        \end{array}
        \\
        \{\} & \textrm{otherwise}
      \end{array}
    \right.
  \end{displaymath}
  Now, for any $z_1,\dots,z_m \in \mathcal{M}$, we know that, for all
  $j$, $(z_j,z_j) \in \rho\ \tyPrimNm{M}\ (e_j)$, so we know that
  $(\tySem{M}z_1\dots z_n, \tySem{M}z_1 \dots z_n) \in \rho\
  \tyPrimNm{M}\ (e)$. Thus there exist $x_1,...,x_n$ such that $e =
  e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free
  monoids, we learn that $x_1,...,x_n$ is the solution we need.
\end{proof}

We can further assume that there are other operations that distribute
over the multiplication of the monoid, for example semi-ring
operations (e.g.~relational algebra).

Examples
\begin{enumerate}
\item Finite sets of stuff.
\item Let $\mathcal{M} = \Sigma^* \times
  \mathcal{P}_{\mathit{fin}}(C)$, where $\Sigma$ is some alphabet, and
  $C$ is some set of taint markers. The centre of this monoid is
  $\{([], c) \sepbar c \in \mathcal{P}_{\mathit{fin}}(C) \}$.
\end{enumerate}

>>>>

\section{Monoid Indexed Types}
\label{sec:monoid-indexed-types}

In the last section we looked at types indexed by abelian groups. We
now examine an instantiation of our general framework with a monoid
type indexed by elements of a commutative monoid. Our main example of
monoid indexed types is strings indexed by ``taint'' markers: when
strings are appended their taint markers are combined. Thanks to the
information in the indexes, we are able to prove an indefinability
result (\thmref{thm:monoid-indefinability}).

We assume a single indexing sort $\mathsf{M}$, with the monoid
operations of unit and multiplication along with the commutative
monoid axioms. There is a single primitive type $\tyPrimNm{M}$, with
$\primTyArity(\tyPrimNm{M}) = [\mathsf{M}]$. For the index-erasure
semantics, we assume some (not necessarily commutative) monoid $M$,
which we also write multiplicatively. Thus, we set
$\tyPrimSem{\tyPrimNm{M}} = M$. The collection of primitive operations
$\Gamma_{\mathit{Mon}}$ is simply the two monoid operations:
\begin{displaymath}
  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{2em}}r@{\hspace{0.5em}:\hspace{0.5em}}l}
    1 & \tyPrim{M}{1}, &
    (\cdot) & \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a b}
  \end{array}
\end{displaymath}
which we interpret using the monoid structure of $M$, giving an
environment $\eta_\Mon \in \ctxtSem{\Gamma_\Mon}$.

For the sets of relational environments, we make use of the
construction in \autoref{sec:constr-rel-env}, setting the chosen model
$\mathcal{M}$ to be the \emph{centre} $Z(M)$ of $M$ (recall that the
centre of a monoid is the set of all elements that commute with every
other element). We then set $R_{\tyPrimNm{M}}(m) = \{ (n,mn) \sepbar n
\in M \}$ and $R^\bullet_{\tyPrimNm{M}} = \emptyset$. By setting the
model $\mathcal{M}$ to only be the elements of $M$ that commute with
everything else, we ensure that the following lemma holds:
\begin{lemma}
  For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
  $(\eta_\Mon,\eta_\Mon) \in \rsem{\Gamma_\Mon}{\relEnv{E}}\rho$.
\end{lemma}

\newcommand{\mscup}{\stackrel{ms}\cup}

\begin{example}
  By selecting a suitable monoid $M$, we can simulate a programming
  language that dynamically annotates strings with taint markers. Let
  $\Sigma$ be a set of characters from some alphabet, and let $T$ be a
  set of taint markers. Set $M = \Sigma^* \times
  \mathcal{MS}_{\mathit{fin}}(T)$, where $\mathcal{MS}_{\mathit{fin}}$
  denotes the finite power set operator. Since both $\Sigma^*$ and
  $\mathcal{MS}_{\mathit{fin}}(T)$ are monoids (the latter with
  multiset union as the monoid multiplication, which we write as
  $\mscup$), $M$ is itself a monoid. The centre of $M$, $Z(M)$,
  consists of all elements of the form $([], t)$ for multisets of
  taint markers.

  As an example of a free theorem for this instantiation of our
  general framework, consider programs of the following type:
  \begin{displaymath}
    ; \Gamma_\Mon \vdash M : \forall a,b\mathord:\mathsf{M}.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{aa}
  \end{displaymath}
  From this type we can derive the following free theorem about
  $M$. For all $a, b \in \mathcal{MS}_{\mathit{fin}}(T)$, $(s_1,t_1),
  (s_2,t_2) \in \Sigma^*$,
  \begin{displaymath}
    \begin{array}{l}
      \pi_2(\tmSem{M}\ (s_1,t_1 \mscup a)\ (s_2,t_2 \mscup b)) = \\
      \hspace{2em}\pi_2(\tmSem{M}\ (s_1,t_1)\ (s_2,t_2)) \mscup a \mscup a
    \end{array}
  \end{displaymath}
  We read this result as follows: if we taint the first input by $a$
  and the second by $b$, then the taint of the output varies according
  to $M$'s type. In this case the output contains two copies of $a$
  and none of $b$.
\end{example}

% if $b$ does not appear in the output, then it cannot have been used in
% the program. Turn this into a theorem...

% If forall i1...,in. M<e0> \to ... \to M<em> -> M<e>
% and e0 is not a factor of e, then we can get rid of the first input

% It ought to be invariant under its first input... Set R^\bullet to
% be the everywhere relation....

% prove this by using a relational environment that excludes

\begin{theorem}
  \begin{equation}
    \label{eq:fo-monoid-type-insensitive}
    ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
  \end{equation}

\end{theorem}

\begin{theorem}
  \label{thm:monoid-indefinability}
  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
  there exists a program $M$ such that
  \begin{displaymath}
    \label{eq:fo-monoid-type}
    ; \Gamma_\Mon \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
  \end{displaymath}
  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e =
  i_1^{b_1}\dots i_m^{b_m}$, if and only if there is a solution
  $\vec{x} = (x_1,\cdots,x_n)$ in the natural numbers to the equation
  $A x^\top = b$.
\end{theorem}

\begin{example}
  By \thmref{thm:monoid-indefinability}, it is impossible to write a
  program with the type $\forall m_1,m_2 \mathord: \mathsf{M}.\
  \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_1}$, showing that it is
  impossible to convert a doubly tainted value to a singly tainted
  one.

  In contrast to the case of abelian group indexed types, the lack of
  inverses in the monoid indexed setting means that the type $\forall
  m_1,m_2\mathord:\mathsf{M}.\ \tyPrim{M}{m_1m_2} \to \tyPrim{M}{m_2}
  \to \tyPrim{M}{m_1}$ is also uninhabited. There are no naturals
  $a_1,a_2$ such that $(m_1m_2)^{a_1}m_2^{a_2} \equiv m_1$.
\end{example}

% \fixme{Finish this} We can further assume that there are other
% operations that distribute over the multiplication of the monoid, for
% example semi-ring operations (e.g.~relational algebra).

\section{Metric Spaces and Simple Continuity Analysis}
\label{sec:continuity-analysis}

\section{Singleton Types}
\label{sec:singleton-types}

\newcommand{\Sing}{\mathit{Sing}}

@@

\begin{enumerate}
\item Given an algebraic theory, and a model, we can define a
  singleton type system.
\item Need to carefully define this in general, and give the
  appropriate relational environments
\item Give the integers (freely generated monoid) as an example
\end{enumerate}

\begin{theorem}
  There exists a program
  \begin{displaymath}
    -; \Gamma_\Sing \vdash M : \forall i_1,...,i_m\mathord:\mathsf{S}.\ \tyPrim{S}{e_1} \to ... \to \tyPrim{S}{e_n} \to \tyPrim{S}{e}
  \end{displaymath}
  if and only if there exists an index expression
  $i'_1\mathord:\mathsf{S}, ..., i'_n\mathord:\mathsf{S} \vdash e' :
  \mathsf{S}$ such that $(e_1,...,e_n)^*e' \equiv e$.
\end{theorem}

\begin{proof}
  (If) Use the index expression $e'$ to define $M$, replacing each
  index operation with its program-level counterpart.

  (Only if) \fixme{use a special relational environment...}
\end{proof}

% \subsection{Predicate Types}
% \label{sec:predicate-types}

% \begin{enumerate}
% \item Let the indexing theory be predicates on the underlying value
% \item With the laws of boolean algebra
% \item Singleton types are a special case (i.e. with equality to a fixed value)
% \item But also have $\land$, $\lor$, $\lnot$ (and derived $\Rightarrow$)
% \item A pair of things is related if they both satisfy the predicate
%   (and are equal?)
% \item Need a special type for comparisons?
%   \begin{displaymath}
%     (<) : \forall z_1,z_2:\mathsf{integer}.\ \tyPrim{int}{z_1} \to \tyPrim{int}{z_2} \to (\tyPrim{int}{< z_1} \times 
%   \end{displaymath}
% \end{enumerate}

\section{Logical Information Flow}
\label{sec:information-flow}

@@

@@
